---
title: "lazyNumbers: exact floating-point arithmetic"
output: github_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/lazyNumbers/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/lazyNumbers/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

It is well-known that floating-point arithmetic is inexact even with some 
simple operations. For example:

```{r}
1 - 7 * 0.1 == 0.3
```

This package provides the *lazy numbers*, which allow exact floating-point 
arithmetic:

```{r}
library(lazyNumbers)
x <- lazynb(1) - lazynb(7) * lazynb(0.1)
as.double(x) == 0.3
```

Here is a more interesting example. Consider the following recursive sequence:

```{r}
u <- function(n) {
  if(n == 1) {
    return(1/7)
  }
  8 * u(n-1) - 1
}
```

It is clear that all terms of this sequence equal `1/7` (approx. `0.1428571`). 
However:

```{r}
u(15)
u(18)
u(20)
u(30)
```

When it is evaluated in double precision, the sequence becomes crazy. This is 
not the case of its lazy version:

```{r}
u <- function(n) {
  if(n == 1) {
    return(1/lazynb(7))
  }
  8 * u(n-1) - 1
}
as.double(u(30))
```


## Blog post

[The lazy numbers in R](https://laustep.github.io/stlahblog/posts/lazyNumbers.html)


## License

This package is provided under the GPL-3 license but it uses the C++ library 
CGAL. If you wish to use CGAL for commercial purposes, you must obtain a 
license from the [GeometryFactory](https://geometryfactory.com).
